Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2018-05-14 articles/formalism/sections/generalizedeb.tex
Further edits through section 3.

%--------------------------------------------------------
%--------------------------------------------------------
\subsubsection{Generalized operators}
In this section we present a systematic method to generalize the real space operators that translate between $Q/U$ and $E/B$ description of CMB polarization.
The azimuthal part of the real space kernels is constrained to be of the form $\sim e^{\pm i2 \alpha}$. The shape of the radial part of the operator is determined by the basis functions and varies as a function of the band limit. It is here that we may potentially choose alternate forms for the radial functions to suit certain kind of analysis.

We can systematically generalize the real space operator by introducing the following harmonic space operator,
%
\beq
\tilde{\mathcal{G}} = {\begin{bmatrix} g_{\ell}^E & 0  \\  0 & g_{\ell}^B \end{bmatrix}} \,,
\eeq
%
where the functions $g_{\ell}^E$ and $g_{\ell}^B$ represent the harmonic representation of the modified radial functions and can in the most general case be chosen to be different for $E$ and $B$ modes. To simplify discussions and without loosing generality we proceed by setting $g_{\ell}^E = g_{\ell}^B= g_{\ell}$. Given this harmonic function $g_{\ell}$, we can define the real space operator $\bar{O}'$ which translates Stokes $Q$ \& $U$ to scalars $E$ \& $B$ and the inverse operator $\bar{O}'^{-1}$ in the following manner,
%
\begin{subequations} \label{eq:gen_qu2eb}
\beqry
{\bar O}' &=& {{}_0\mathcal{Y}} *\tilde T^{-1}*\tilde{\mathcal{G}}* {{}_2\mathcal{Y}^{\dagger}} *\bar T \,,\\
{\bar O}'^{-1}&=& \bar{T}^{-1} *{{}_2\mathcal{Y}}* \tilde{\mathcal{G}}^{-1} *\tilde T *{{}_0\mathcal{Y}^{\dagger}}
\eeqry
\end{subequations}
%
where we have used the primed notation  to distinguish these generalized operators from the default operators defined in \sec{sec:qu2eb} and \sec{sec:eb2qu}. Requiring both the forward and inverse operators to be well defined constrains the choice of $\tilde{\mathcal{G}}$ to be such that it has a valid  inverse. This is an important criteria to recover the standard CMB spectra. The radial parts of this generalized operator and its inverse are given by the following expressions,
%
\begin{subequations}\label{eq:generalized_radial_kernel}
\beqry 
G_{QU \rightarrow EB}(\beta) &=& G(\beta) = \sum _{\ell=2} ^{\ell_{\rm max}} g_{\ell}\frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell + 2)!}} P_{\ell}^2(\cos{\beta}) \, \label{eq:mod_rad_forward} \,, \\
G_{EB \rightarrow QU}(\beta) &=& G^{-1}(\beta) = \sum _{\ell=2} ^{\ell_{\rm max}} g_{\ell}^{-1}\frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell + 2)!}} P_{\ell}^2(\cos{\beta}) \,. \label{eq:mod_rad_inverse}
\eeqry
\end{subequations}
%
The default radial function is just a special case resulting from the choice $\tilde{\mathcal{G}}=\mathbb{1}$ ($g_{\ell}=1$), in which case $\tilde{\mathcal{G}}^{-1}=\tilde{\mathcal{G}}$ and therefore $G^{-1}(\beta) = G(\beta)={{}_{\mm}f}$.

While defining these generalized operators, it is more natural to choose the real space function $G(\beta)$, bearing in mind the constraint that $G(\beta=0)=G(\beta=\pi)=0$. Employing the orthogonality property of associated Legendre polynomials it can be shown that the harmonic function $g_{\ell}$ is given by the expression,
%
\beq
g_{\ell} = 2 \pi \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} \int _{0}^{\pi} G(\beta) P_{\ell}^{2}(\cos{\beta}) d\cos{\beta} \,. \label{eq:gb2bl}
\eeq
%

An arbirtrary $G(\beta)$ for which $g_{\ell} \neq 1$ can be equivalently thought in terms of the standard E and B fields being convolved with some effective circularly symmetric instrument beam whose radial profile is given by the expression,
%
\beq
B(\beta) = \sum_{\ell=0}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} g_{\ell} P_{\ell}^{0} (\cos{\beta})\,,
\eeq
%
where $g_{\ell}$ is the same harmonic function as that appearing in \eq{eq:generalized_radial_kernel}.
In contrast to the radial function $G(\beta)$ an instrumental beam function appropriately normalized has the property $B(\beta) \rightarrow 1$ as $\beta \rightarrow 0$. Though the real space behavior of these two function $G(\beta)$ and $B(\beta)$ has these differences, in harmonic space they play identical roles.
%
\begin{figure}[!t] 
\centering
\subfigure[\label{fig:bl_gbeta}]{\includegraphics[width=0.48\columnwidth]{Gbeta_for_different_gl_lmax1536.pdf}}
\subfigure[\label{fig:glbl}]{\includegraphics[width=0.48\columnwidth]{gl_for_different_gbeta_lmax1536.pdf}}
\subfigure[\label{fig:gl_bbeta}]{\includegraphics[width=0.48\columnwidth]{Bbeta_for_gl_lmax1536.pdf}}
\caption{\textit{Top left:} The green line depicts the default radial kernel $f(\beta)$ defined in \eq{eq:qu2eb_gen_kernel}, multiplied by an apodized step function $\theta(3 \beta_0)$. The blue and orange lines depict the modified radial function resulting the beam harmonics $b_{\ell}$ corresponding to Gaussian beams with fwhm=15 \& 12 arc-minutes respectively. The red curve depicts an example modified radial function: $G(\beta)=\mathcal{A} \beta^{n_o} \exp{\left[ -\left( \frac{\beta-\beta_s}{\sqrt{2} \sigma} \right)^{n_1} \right]}$ with parameters set to the following values $[n_0=1;\, \beta_s=0 ;\, \sigma = 0.004 ;\, n_1=1.5]$. The black dashed curve depicts the band limited reconstruction of the modified radial function. \textit{Top right:} This figure depicts the harmonic representation of the respective radial functions as indicated by the legend. The dashed curves of the corresponding color depict the inverse of the harmonic functions. \textit{Bottom:} This figure depicts the normalized beam function $B(\beta)$ evaluated from interpreting the respective harmonic functions as those corresponding to an effective instrument beam.}
\label{fig:example_gbeta}
\end{figure}
%

 Therefore it is possible to interpret the beam harmonic coefficients as those representing some modified radial kernel. \fig{fig:glbl} depicts the harmonic functions $g_{\ell}$ for the respective radial kernel and beams.  The modified radial kernel resulting from Gaussian beams with ${\rm fwhm} =15' \,\&\, 12'$ are depicted in \fig{fig:bl_gbeta} as blue and orange curves respectively. Note that instruments  beams tend to increase the non-locality parameter $\beta_0$, indicated by the shifting right of the maxima of the respective kernels, as one may have expected. The red curve depicts a modified radial kernel which by construction has a very small $\beta_0$.  Note that the beam function corresponding to the default radial kernel ($g_{\ell}=1$) is merely a band limited representation of the delta function beam. \fig{fig:gl_bbeta} depicts the radial profile of the  effective beams corresponding to the different radial kernels.

%In \sec{sec:local_conv_eb} we constructed localized convolution kernels by multiplying $R(\beta)$ with an apodized version of the step function $\theta_{\rm apo}(r_{\rm cutoff})$. The oscillation seen in the spectra in \fig{fig:eb-spectra_rad_cutoff} can be explained to be due to this effective beam characterized by $g_{\ell}^2$ operating on the power spectra. The effective beam can be evaluated by computing the multipole function $g_{\ell}$ as follows,
%%
%\beq
%g_{\ell} = 2 \pi \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} \int _{0}^{r_{\rm cutoff}} R(\beta) \theta_{\rm apo}(r_{\rm cutoff})  P_{\ell}^{2}(\cos{\beta}) d\cos{\beta} \,, \label{eq:gb2bl} \,,
%\eeq
%%
%where the upper limit of the integration is set to $r_{\rm cutoff}$ since the function $\theta_{\rm apo}(r_{\rm cutoff})$ vanishes for $\beta>r_{\rm cutoff}$. The function $b^2_{\ell}-1$ matches the oscillation seen in \fig{fig:eb-spectra_rad_cutoff} as  seen in \fig{fig:match_cl_oscillations} where the two results have been over plotted.
%%
%\begin{figure}[!t] 
%\centering
%\subfigure[]{\includegraphics[width=0.98\columnwidth]{analytical_cl_oscillations_vs_data.pdf}}
%\caption{The thin lines depicts the same spectral differences as those seen in \fig{fig:eb-spectra_rad_cutoff}, while the thick lines of the corresponding color depict the function $g_{\ell}^2 -1$ as derived from evaluating \eq{eq:gb2bl} for different $r_{\rm cutoff}$}.
%\label{fig:match_cl_oscillations}
%\end{figure}
%%
%The apodized step function in this case transition from 1 at $\beta < r_{\rm cutoff} -w$ to 0 at $r_{\rm cutoff}$ over a width $w= 3^{\circ}$ with a cosine squared profile .

%--------------------------------------------------------
\subsubsection{Recovering the default $E$ and $B$ mode spectra}
The generalized convolution kernels defined in the previous section, when operated on the Stokes vector returns some scalar $E'$ and $B'$ mode maps,
%
\beq
\bar{S}' = \bar{O}' * \bar{P}
\eeq
%
which are merely filtered version of he standard $E$ and $B$ modes maps. Since the harmonic functions $g_{\ell}$ of the modified radial function $G(\beta)$ can be simply interpreted as the harmonic coefficients of some azimuthally symmetric beam, the spectra of the scalar fields $E'$ and $B'$ are related to the spectra of the standard $E$ and $B$ fields via the following relation, 
 %
 \begin{subequations}
 \beqry
C_{\ell}^{EE,BB,EB} &= &C_{\ell}^{E'E',B'B',E'B'} /   g_{\ell}^2\,,\\
C_{\ell}^{TE,TB}  &=&  C_{\ell}^{TE',TB'} / g_{\ell}\,,
 \eeqry
 \end{subequations}
 %
 where $C_{\ell}$ denotes the angular power spectra and $T$ refers to the temperature anisotropy map. Therefore the standard CMB spectra can be always recovered as long as the $1/g_{\ell}$ \& $1/g_{\ell}^2$ are well behaved functions, which can be ensured by making a suitable choice for the modified radial function $G(\beta)$. Some examples forms of $G(\beta)$, its harmonic representation $g_{\ell}$ (and its inverse $1/g_{\ell}$ ) and the corresponding beam $B(\beta)$ are depicted in \fig{fig:example_gbeta}.
%One direct application of this freedom in choosing the radial kernel is that one can mitigate the issue of leakage of power from $E$ to $B$.  By constructing a radial function which tapers to zero at sufficiently small radii one can clearly discard pixels which are affected by masking. 
%--------------------------------------------------------

>>>>

%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Generalized polarization operators}
With our better understand of the radial part of the kernel for CMB polarization, we can write down generalized $E/B$-like fields that depend on a different radial function, even one that we specify to have compact support.
The spin symmetry constrains the the azimuthal part of the real space kernels is constrained to be of the form $\sim e^{\pm i2 \alpha}$.  The shape of the radial part of the operator is determined by the basis functions and varies as a function of the band limit. It is here that we may potentially choose alternate forms for the radial functions to suit certain kind of analysis.

We can systematically generalize the real space operator by introducing the following harmonic space filter function,
%
\beq
\tilde{\mathcal{G}} = {\begin{bmatrix} g_{\ell}^E & 0  \\  0 & g_{\ell}^B \end{bmatrix}} \,,
\eeq
%
where the functions $g_{\ell}^E$ and $g_{\ell}^B$ represent the harmonic representation of the modified radial functions and can in the most general case be chosen to be different for $E$ and $B$ modes. To simplify discussions, we proceed by setting $g_{\ell}^E = g_{\ell}^B= g_{\ell}$. Given this harmonic function $g_{\ell}$, we can define the real space operator $\bar{O}'$ which translates Stokes $Q/U$ to scalars $E/B$ and the inverse operator $\bar{O}'^{-1}$ in the following manner,
%
\begin{subequations} \label{eq:gen_qu2eb}
\beqry
{\bar O}' &=& {{}_0\mathcal{Y}} \, \tilde T^{-1} \tilde{\mathcal{G}} {{}_2\mathcal{Y}^{\dagger}} \, \bar T \,,\\
{\bar O}'^{-1}&=& \bar{T}^{-1} {{}_2\mathcal{Y}}\, \tilde{\mathcal{G}}^{-1} \tilde T {{}_0\mathcal{Y}^{\dagger}}.
\eeqry
\end{subequations}
%
The primed notation distinguishes these generalized operators from the default operators defined in \sec{sec:qu2eb} and \sec{sec:eb2qu}. We require both the forward and inverse operators to be well defined.   This constrains the choice of $\tilde{\mathcal{G}}$ to have a valid  inverse and is important to recover the standard CMB power spectra. The radial parts of this generalized operator and its inverse are given by the following expressions,
%
\begin{subequations}\label{eq:generalized_radial_kernel}
\beqry 
G_{QU \rightarrow EB}(\beta) &=& G(\beta) = \sum _{\ell=2} ^{\ell_{\rm max}} g_{\ell}\frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell + 2)!}} P_{\ell}^2(\cos{\beta}) \, \label{eq:mod_rad_forward} \,, \\
G_{EB \rightarrow QU}(\beta) &=& G^{-1}(\beta) = \sum _{\ell=2} ^{\ell_{\rm max}} g_{\ell}^{-1}\frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell + 2)!}} P_{\ell}^2(\cos{\beta}) \,. \label{eq:mod_rad_inverse}
\eeqry
\end{subequations}
%
The default radial function is just a special case resulting from the choice $\tilde{\mathcal{G}}=\mathbb{1}$ ($g_{\ell}=1$), in which case $\tilde{\mathcal{G}}^{-1}=\tilde{\mathcal{G}}$ and therefore $G^{-1}(\beta) = G(\beta)={{}_{\mm}f}$.

While defining these generalized operators, it is more natural to choose the real space function $G(\beta)$ rather than the harmonic space $g_l$, bearing in mind the constraint that $G(\beta=0)=G(\beta=\pi)=0$. Employing the orthogonality property of associated Legendre polynomials it can be shown that the harmonic function $g_{\ell}$ is given by the expression,
%
\beq
g_{\ell} = 2 \pi \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} \int _{0}^{\pi} G(\beta) P_{\ell}^{2}(\cos{\beta}) d\cos{\beta} \,. \label{eq:gb2bl}
\eeq
%

An arbirtrary $G(\beta)$ for which $g_{\ell} \neq 1$ can be equivalently thought in terms of the standard $E/B$ fields being convolved with some effective circularly symmetric beam whose radial profile is given by the expression,
%
\beq
B(\beta) = \sum_{\ell=0}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} g_{\ell} P_{\ell}^{0} (\cos{\beta})\,,
\eeq
%
where $g_{\ell}$ is the same harmonic function as that appearing in \eq{eq:generalized_radial_kernel}.  \revisit{Need another symbol than $B$ for beam.}
In contrast to the radial function $G(\beta)$ an instrumental beam function appropriately normalized has the property $B(\beta) \rightarrow 1$ as $\beta \rightarrow 0$. Though the real space behavior of these two function $G(\beta)$ and $B(\beta)$ has these differences, in harmonic space they play identical roles.
%
\begin{figure}[!t] 
\centering
\subfigure[\label{fig:bl_gbeta}]{\includegraphics[width=0.48\columnwidth]{Gbeta_for_different_gl_lmax1536.pdf}}
\subfigure[\label{fig:glbl}]{\includegraphics[width=0.48\columnwidth]{gl_for_different_gbeta_lmax1536.pdf}}
\subfigure[\label{fig:gl_bbeta}]{\includegraphics[width=0.48\columnwidth]{Bbeta_for_gl_lmax1536.pdf}}
\caption{\textit{Top left:} The green line depicts the default radial kernel $f(\beta)$ defined in \eq{eq:qu2eb_gen_kernel}, multiplied by an apodized step function $\theta(3 \beta_0)$. The blue and orange lines depict the modified radial function resulting the beam harmonics $b_{\ell}$ corresponding to Gaussian beams with fwhm=15 \& 12 arc-minutes respectively. The red curve depicts an example modified radial function: $G(\beta)=\mathcal{A} \beta^{n_o} \exp{\left[ -\left( \frac{\beta-\beta_s}{\sqrt{2} \sigma} \right)^{n_1} \right]}$ with parameters set to the following values $[n_0=1;\, \beta_s=0 ;\, \sigma = 0.004 ;\, n_1=1.5]$. The black dashed curve depicts the band limited reconstruction of the modified radial function. \textit{Top right:} This figure depicts the harmonic representation of the respective radial functions as indicated by the legend. The dashed curves of the corresponding color depict the inverse of the harmonic functions. \textit{Bottom:} This figure depicts the normalized beam function $B(\beta)$ evaluated from interpreting the respective harmonic functions as those corresponding to an effective instrument beam.}
\label{fig:example_gbeta}
\end{figure}
%

% Therefore it is possible to interpret the beam harmonic coefficients as those representing some modified radial kernel.
In \fig{fig:example_gbeta}, we examine in more detail the relationship between the modified radial kernels and these beam harmonic coefficients.  
 \fig{fig:gl_bbeta} depicts the radial profile of the  effective beams corresponding to different radial kernels: the standard kernel with a radial cutoff, kernels corresponding to Gaussian smoothings of the $E/B$ fields, an radial function without oscillations and an exponential cutoff.  Note that the smoothing tend to increase the non-locality, indicated by the shifting right of the maxima of the respective kernels, as one may have expected.  The exponential cutoff (red curve) by construction has a very small non-locality (parameterized by $\beta_0$).

\fig{fig:glbl} depicts the harmonic description $g_{\ell}$ for these respective radial kernels and beams.
 %The modified radial kernel resulting from Gaussian beams with ${\rm fwhm} =15' \,\&\, 12'$ are depicted in \fig{fig:bl_gbeta} as blue and orange curves respectively.
Finally, the beam that the modified radial kernels appies to the $E/B$ fields we show in \fig{fig:gl_bbeta}.  Note that the beam function corresponding to the default radial kernel ($g_{\ell}=1$) is merely a band limited representation of a delta-function beam.
%In \sec{sec:local_conv_eb} we constructed localized convolution kernels by multiplying $R(\beta)$ with an apodized version of the step function $\theta_{\rm apo}(r_{\rm cutoff})$. The oscillation seen in the spectra in \fig{fig:eb-spectra_rad_cutoff} can be explained to be due to this effective beam characterized by $g_{\ell}^2$ operating on the power spectra. The effective beam can be evaluated by computing the multipole function $g_{\ell}$ as follows,
%%
%\beq
%g_{\ell} = 2 \pi \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} \int _{0}^{r_{\rm cutoff}} R(\beta) \theta_{\rm apo}(r_{\rm cutoff})  P_{\ell}^{2}(\cos{\beta}) d\cos{\beta} \,, \label{eq:gb2bl} \,,
%\eeq
%%
%where the upper limit of the integration is set to $r_{\rm cutoff}$ since the function $\theta_{\rm apo}(r_{\rm cutoff})$ vanishes for $\beta>r_{\rm cutoff}$. The function $b^2_{\ell}-1$ matches the oscillation seen in \fig{fig:eb-spectra_rad_cutoff} as  seen in \fig{fig:match_cl_oscillations} where the two results have been over plotted.
%%
%\begin{figure}[!t] 
%\centering
%\subfigure[]{\includegraphics[width=0.98\columnwidth]{analytical_cl_oscillations_vs_data.pdf}}
%\caption{The thin lines depicts the same spectral differences as those seen in \fig{fig:eb-spectra_rad_cutoff}, while the thick lines of the corresponding color depict the function $g_{\ell}^2 -1$ as derived from evaluating \eq{eq:gb2bl} for different $r_{\rm cutoff}$}.
%\label{fig:match_cl_oscillations}
%\end{figure}
%%
%The apodized step function in this case transition from 1 at $\beta < r_{\rm cutoff} -w$ to 0 at $r_{\rm cutoff}$ over a width $w= 3^{\circ}$ with a cosine squared profile .

%--------------------------------------------------------
\subsubsection{Recovering the standard $E$ and $B$ mode spectra}
The generalized convolution kernels defined in the previous section, when operated on the Stokes vector returns some scalar $E'$ and $B'$ mode maps,
%
\beq
\bar{S}' = \bar{O}' \bar{P}
\eeq
%
which are merely filtered versions of the standard $E/B$ modes maps. Since the filter function is simply $g_{\ell}$, easily obtained from the modified radial function $G(\beta)$, and it can be simply interpreted as the harmonic coefficients of some azimuthally symmetric beam, the power spectra of the modified scalar fields $E'$ and $B'$ are related to the spectra of the standard $E$ and $B$ fields via the following relation, 
 %
 \begin{subequations}
 \beqry
C_{\ell}^{EE,BB,EB} &= &C_{\ell}^{E'E',B'B',E'B'} /   g_{\ell}^2\,,\\
C_{\ell}^{TE,TB}  &=&  C_{\ell}^{TE',TB'} / g_{\ell}\,,
 \eeqry
 \end{subequations}
 %
 where $C_{\ell}$ denotes the angular power spectra and $T$ refers to the temperature anisotropy map. Therefore the standard CMB spectra can always be recovered as long as the $1/g_{\ell}$ and $1/g_{\ell}^2$ are well behaved functions, which can be ensured by making a suitable choice for the modified radial function $G(\beta)$. %Some examples forms of $G(\beta)$, its harmonic representation $g_{\ell}$ (and its inverse $1/g_{\ell}$ ) and the corresponding beam $B(\beta)$ are depicted in \fig{fig:example_gbeta}.
%One direct application of this freedom in choosing the radial kernel is that one can mitigate the issue of leakage of power from $E$ to $B$.  By constructing a radial function which tapers to zero at sufficiently small radii one can clearly discard pixels which are affected by masking. 
%--------------------------------------------------------
